Capstone Movielens Report

### 2025-02-09 reports/capstone-movielens-report.Rmd
Report: Simplest (naive) model described.

Let us install the development version of this package from the GitHub repository and attach the correspondent library to the global environment:
```{r }
if(!require(edx.capstone.movielens.data)) pak::pak("AzKurban-edX-DS/edx.capstone.movielens.data")

@@

summary(edx)
summary(final_holdout_test)
```

Let us look into the details of the `edx` dataset:
``` {r echo = TRUE}
str(edx)
```
Note that we have 9000055 rows and six columns in there:  
```{r }
dim_edx <- dim(edx)
print(dim_edx)
```

@@

Next, we have 10677 different movies 
```{r}
n_movies <- n_distinct(edx$movieId)
print(n_movies)
```
and 69878 different users in the dataset:
```{r}
n_users <- n_distinct(edx$userId)
print(n_users)
```

Now, note the expressions below which confirm the fact explained in [Section 23.1.1 Movielens data](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_[@IDS2]:
```{r}
max_possible_ratings <- n_movies*n_users
sprintf("Maximum possible ratings: %s", max_possible_ratings)
sprintf("Rows in `edx` dataset: %s", dim_edx[1])
sprintf("Not every movie was rated: %s", max_possible_ratings > dim_edx[1])

The following plot of the matrix for a random sample of 100 movies and 100 users with yellow indicating a user/movie combination for which we have a rating shows how _sparse_ the matrix is[@IDS2_23-1-1]:
```{r sparsity-of-movie-recs, echo=TRUE, fig.width=3, fig.height=3, out.width="40%"}
users <- sample(unique(edx$userId), 100)

Taking into consideration these observations, we came up with a decision to use the data from the dataset only for users who have provided at least 100 ratings[@IDS2_23-1-1].

#### Movie genres
\
The following code computes movie rating summaries by popular genres like Drama, Comedy, Thriller, and Romance:
```{r}
#library(stringr)
genres = c("Drama", "Comedy", "Thriller", "Romance")
sapply(genres, function(g) {
  sum(str_detect(edx$genres, g))
})
```

@@

#### Rating distribution plot[@MRS-R-BEST]
\
The code below demonstrates another way of visualizing the rating distribution:
```{r}
edx |>
  group_by(rating) |>
  summarize(count = n()) |>
  ggplot(aes(x = rating, y = count)) +
  geom_bar(stat = "identity", fill = "#8888ff") +
  ggtitle("Rating Distribution") +
  xlab("Rating") +
  ylab("Occurrences Count") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))

@@

#### Yearly rating count[@MRS-R-BEST]
\
```{r}
print(edx |> 
  mutate(year = year(as_datetime(timestamp, origin = "1970-01-01"))) |>
  group_by(year) |>
  summarize(count = n())
)
```

#### Average rating per year plot[@MRS-R-BEST]
\
```{r}
edx |> 
  mutate(year = year(as_datetime(timestamp, origin = "1970-01-01"))) |>
  group_by(year) |>
  summarize(avg = mean(rating)) |>
  ggplot(aes(x = year, y = avg)) +
  geom_bar(stat = "identity", fill = "#8888ff") + 
  ggtitle("Average rating per year") +
  xlab("Year") +
  ylab("Average rating") +
  scale_y_continuous(labels = comma) + 
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

#### Movie popularity count[@MRS-R-BEST]
\
```{r}
print(edx |> 
  group_by(movieId) |> 
  summarize(count = n()) |>
  slice_head(n = 10)
)
```

@@

#### Ratings per movie plot[@MRS-R-BEST]
\
```{r}
edx |>
  group_by(movieId) |>
  summarize(count = n()) |>
  ggplot(aes(x = movieId, y = count)) +
  geom_point(alpha = 0.2, color = "#4020dd") +
  geom_smooth(color = "red") +
  ggtitle("Ratings per movie") +
  xlab("Movies") +
  ylab("Number of ratings") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

#### Movies' rating histogram[@MRS-R-BEST]
\
```{r}
edx |>
  group_by(movieId) |>
  summarize(count = n()) |>
  ggplot(aes(x = count)) +
  geom_histogram(fill = "#8888ff", color = "#4020dd") +
  ggtitle("Movies' rating histogram") +
  xlab("Rating count") +
  ylab("Number of movies") +
  scale_y_continuous(labels = comma) +
  scale_x_log10(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

#### Ratings per user[@MRS-R-BEST]

##### User rating count (activity measure)
\
```{r}
print(edx |> 
  group_by(userId) |> 
  summarize(count = n()) |>
  slice_head(n = 10)
)
```

##### User rating summary
\
```{r}
summary(edx |> group_by(userId) |> summarize(count = n()) |> select(count))
```

##### Ratings per user plot
\
```{r}
edx |>
  group_by(userId) |>
  summarize(count = n()) |>
  ggplot(aes(x = userId, y = count)) +
  geom_point(alpha = 0.2, color = "#4020dd") +
  geom_smooth(color = "red") +
  ggtitle("Ratings per user") +
  xlab("Users") +
  ylab("Number of ratings") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

##### Users' rating histogram
\
```{r}
edx |>
  group_by(userId) |>
  summarize(count = n()) |>
  ggplot(aes(x = count)) +
  geom_histogram(fill = "#8888ff", color = "#4020dd") +
  ggtitle("Users' rating histogram") +
  xlab("Rating count") +
  ylab("Number of users") +
  scale_y_continuous(labels = comma) +
  scale_x_log10(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

@@

@@

@@

@@

#### [@MRS-R-BEST]
\
```{r}

## Methods / Analysis

You can also embed plots, for example:

```{r pressure, echo=FALSE}
#plot(pressure)
```

>>>>

Let's install the development version of this package from the GitHub repository and attach the correspondent library to the global environment:
```{r }
if(!require(edx.capstone.movielens.data)) pak::pak("AzKurban-edX-DS/edx.capstone.movielens.data")

@@

summary(edx)
summary(final_holdout_test)
```
#### `edx` Dataset
\
Let's look into the details of the `edx` dataset:
``` {r echo = TRUE}
str(edx)
```
Note that we have 9000055 rows and six columns in there:  
```{r }
dim_edx <- dim(edx)
print(dim_edx)
```

@@

##### Movie Genres Data 
\
The following code computes movie rating summaries by popular genres like Drama, Comedy, Thriller, and Romance:
```{r}
#library(stringr)
genres = c("Drama", "Comedy", "Thriller", "Romance")
sapply(genres, function(g) {
  sum(str_detect(edx$genres, g))
})
```

@@

##### Rating distribution plot[@MRS-R-BEST]
\
The code below demonstrates another way of visualizing the rating distribution:
```{r}
edx |>
  group_by(rating) |>
  summarize(count = n()) |>
  ggplot(aes(x = rating, y = count)) +
  geom_bar(stat = "identity", fill = "#8888ff") +
  ggtitle("Rating Distribution") +
  xlab("Rating") +
  ylab("Occurrences Count") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))

@@

##### Yearly rating count[@MRS-R-BEST]
\
```{r}
print(edx |> 
  mutate(year = year(as_datetime(timestamp, origin = "1970-01-01"))) |>
  group_by(year) |>
  summarize(count = n())
)
```

##### Average rating per year plot[@MRS-R-BEST]
\
```{r}
edx |> 
  mutate(year = year(as_datetime(timestamp, origin = "1970-01-01"))) |>
  group_by(year) |>
  summarize(avg = mean(rating)) |>
  ggplot(aes(x = year, y = avg)) +
  geom_bar(stat = "identity", fill = "#8888ff") + 
  ggtitle("Average rating per year") +
  xlab("Year") +
  ylab("Average rating") +
  scale_y_continuous(labels = comma) + 
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

##### Ratings per movie
\

###### Movie popularity count[@MRS-R-BEST]
\
```{r}
print(edx |> 
  group_by(movieId) |> 
  summarize(count = n()) |>
  slice_head(n = 10)
)
```

@@

###### Ratings per movie plot[@MRS-R-BEST]
\
```{r}
edx |>
  group_by(movieId) |>
  summarize(count = n()) |>
  ggplot(aes(x = movieId, y = count)) +
  geom_point(alpha = 0.2, color = "#4020dd") +
  geom_smooth(color = "red") +
  ggtitle("Ratings per movie") +
  xlab("Movies") +
  ylab("Number of ratings") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

###### Movies' rating histogram[@MRS-R-BEST]
\
```{r}
edx |>
  group_by(movieId) |>
  summarize(count = n()) |>
  ggplot(aes(x = count)) +
  geom_histogram(fill = "#8888ff", color = "#4020dd") +
  ggtitle("Movies' rating histogram") +
  xlab("Rating count") +
  ylab("Number of movies") +
  scale_y_continuous(labels = comma) +
  scale_x_log10(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

##### Ratings per user[@MRS-R-BEST]
\

###### User rating count (activity measure)
\
```{r}
print(edx |> 
  group_by(userId) |> 
  summarize(count = n()) |>
  slice_head(n = 10)
)
```

###### User rating summary
\
```{r}
summary(edx |> group_by(userId) |> summarize(count = n()) |> select(count))
```

###### Ratings per user plot
\
```{r}
edx |>
  group_by(userId) |>
  summarize(count = n()) |>
  ggplot(aes(x = userId, y = count)) +
  geom_point(alpha = 0.2, color = "#4020dd") +
  geom_smooth(color = "red") +
  ggtitle("Ratings per user") +
  xlab("Users") +
  ylab("Number of ratings") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

###### Users' rating histogram
\
```{r}
edx |>
  group_by(userId) |>
  summarize(count = n()) |>
  ggplot(aes(x = count)) +
  geom_histogram(fill = "#8888ff", color = "#4020dd") +
  ggtitle("Users' rating histogram") +
  xlab("Rating count") +
  ylab("Number of users") +
  scale_y_continuous(labels = comma) +
  scale_x_log10(n.breaks = 10) +
  theme_economist() +
  theme(axis.title.x = element_text(vjust = -5, face = "bold"), 
        axis.title.y = element_text(vjust = 10, face = "bold"), 
        plot.margin = margin(0.7, 0.5, 1, 1.2, "cm"))
```

@@

\
```{r echo=FALSE}
# #### [@MRS-R-BEST]
```

## Methods / Analysis
### Defining helper functions
\

Let's define some helper functions that we will use in our subsequent analysis:
```{r}
start_date <- function(){
  print(date())
  Sys.time()
}
end_date <- function(start){
  print(date())
  Sys.time() - start
}
rmse <- function(r) sqrt(mean(r^2))
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

### Preparing train and set datasets
\
First, let's note that we have 10677 different movies: 
```{r}
n_movies <- n_distinct(edx$movieId)
print(n_movies)
```
and 69878 different users in the dataset:
```{r}
n_users <- n_distinct(edx$userId)
print(n_users)
```

Now, note the expressions below which confirm the fact explained in [Section 23.1.1 Movielens data](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_[@IDS2] that not every user rated every movie:
```{r}
max_possible_ratings <- n_movies*n_users
sprintf("Maximum possible ratings: %s", max_possible_ratings)
sprintf("Rows in `edx` dataset: %s", dim_edx[1])
sprintf("Not every movie was rated: %s", max_possible_ratings > dim_edx[1])

@@

The following plot of the matrix for a random sample of 100 movies and 100 users with yellow indicating a user/movie combination for which we have a rating shows how _sparse_ the matrix is:
```{r sparsity-of-movie-recs, echo=TRUE, fig.width=3, fig.height=3, out.width="40%"}
users <- sample(unique(edx$userId), 100)

Taking into consideration these observations, we came up with a decision to use the data from the dataset only for users who have provided at least 100 ratings.

Now let's split the `edx` dataset into a training set, which we will use to build and train our models, and a test set in which we will compute the accuracy of our predictions, the way described in the [Section 23.1.1 Movielens data](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_ mentioned above[@IDS2_23-1-1]:
```{r}
# Let's ignore the data for users who have not provided at least 100 ratings:
edx100 <- edx |> 
  group_by(userId) |>
  filter(n() >= 100) |>
  ungroup()

print(edx100 |> summarize(n_distinct(userId), n_distinct(movieId)))

# For each one of these users, we will split their ratings into 80% for training 
# and 20% for testing:

set.seed(2006)
indexes <- split(1:nrow(edx100), edx100$userId)
test_ind <- sapply(indexes, function(i) sample(i, ceiling(length(i)*.2))) |> 
  unlist() |>
  sort()

test_set <- edx100[test_ind,] 
train_set <- edx100[-test_ind,]

# To make sure we don’t include movies in the training set that should not be 
# there, we remove entries using the semi_join function:
test_set <- test_set |> semi_join(train_set, by = "movieId") |> as.data.frame()
summary(test_set)

train_set <- mutate(train_set, userId = factor(userId), movieId = factor(movieId))
summary(train_set)
```

We will use the array representation described in [Section 17.5 of the Textbook](https://rafalab.dfci.harvard.edu/dsbook-part-2/linear-models/treatment-effect-models.html#sec-anova), for the training data: we denote ranking for movie $j$ by user $i$ as $y_{i,j}$. To create this matrix, we use `tidyr::pivot_wider` function:

```{r}
y <- dplyr::select(train_set, movieId, userId, rating) |>
pivot_wider(names_from = movieId, values_from = rating) |>
column_to_rownames("userId") |>
as.matrix()

dim(y)
```

To be able to map movie IDs to titles we create the following lookup table:
```{r}
movie_map <- train_set |> dplyr::select(movieId, title, genres) |> 
  distinct(movieId, .keep_all = TRUE)

summary(movie_map)
```

Note that titles cannot be considered unique, so we can't use them as IDs[@IDS2_23-1-1].

### Naive Model
\
Let's begin our analysis by evaluating the simplest model described in [Section 23.3 The First Model of the *Course Textbook*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#a-first-model), and then gradually refine it through further research.
It is about a model that assumes the same rating for all movies and users with all the differences explained by random variation would look as follows:

$$
Y_{i,j} = \mu + \varepsilon_{i,j}
$$

with $\varepsilon_{i,j}$ independent errors sampled from the same distribution centered at 0 and $\mu$ the _true_ rating for all movies.

We know that the estimate that minimizes the RMSE is the least squares estimate of $\mu$ and, in this case, is the average of all ratings:

```{r pressure, echo=TRUE}
mu <- mean(y, na.rm = TRUE)
print(mu)
```

If we predict all unknown ratings with $\hat{\mu}$, we obtain the following RMSE:

```{r}
rmse(test_set$rating - mu)
```

If we plug in any other number, we will get a higher RMSE. Let's prove that by the following small investigation:
```{r}
plugged_numbers <- seq(0, 6, 0.1) - 3
print(plugged_numbers)

rmses <- sapply(plugged_numbers, function(x){
  rmse(test_set$rating - mu + x)
})

plot(plugged_numbers, rmses, type = "l")

sprintf("Minimum RMSE is achieved when `plugged number is`: %s", plugged_numbers[which.min(rmses)])

@@

To win the grand prize of $1,000,000, a participating team had to get an RMSE of at least 0.8563[@BigChaosSln]. So we can definitely do better![@IDS2_23-3]

### Taking into account User effects
\

@@
